Alternative Methods for Retaining Explicit and Finding Implicit Sharing in Embedded DSLs

### 2023-02-14 paper/paper.tex
enlarged figures

In the toy example
\begin{minted}{haskell}
class Exp repr where
  variable :: String -> repr Int
  constant :: String -> repr Int
  add :: repr Int -> repr Int -> repr Int
  novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}
the function \mintinline{haskell}{novel} will exhibit an issue in sharing of
it's two outputs when writing it's implementation that cannot be solved using
the methods of \cite{kiselyov:sharing}. We will illustrate this implementation
issue in section \ref{limitexplicit}. In our work, this translated into the
inability to process large library functions.

@@

We used this language to implement approximations of vector \texttt{cos} and \texttt{tan} (i.e., a
loop body that computes over an input and output array). The loop body was
unrolled by a factor of four making for a sufficiently large resulting DAG.

We profiled cos using both Kiselyov's method (i.e., hash-consing) and are own
(i.e., trie-consing with some explicit sharing). It wasn't possible to use
Kiselyov's explicit sharing method where meaningful so our code was able to
achieve greater performance, with only a small increase in memory consumption
(see figure \ref{fig:hashvscons}).

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/cos_profile_hashcons.png}
  \hfill
  \includegraphics[width=0.45\textwidth]{figs/cos_profile_triecons.png}
  \caption{Comparison of cos using hash-cons vs trie-cons}
  \label{fig:hashvscons}
\end{figure}

For tan, we were unable to generate a resulting code graph using Kiselyov's
method. Using explicit sharing via let constructs wasn't possible due to the
multiple outputs issue and without it the amount of computation due to redundant
hash-consing was simply too large to perform in a reasonable amount of time.

We were able to generate tan using our method both without and with our explicit
sharing, however without explicit sharing we consumed an unreasonable amount of
memory (see figure \ref{fig:explicitvsno}). It should be noted, the main source
of redundant computation in tan is the reuse of computed values of cos and sin.
By simply explicitly sharing just those values we achieved the significant
speedup shown in figure \ref{fig:explicitvsno}.

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/noexplicit.png}
  \hfill
  \includegraphics[width=0.45\textwidth]{figs/explicit.png}
  \caption{Comparison of tan profiling using no explicit sharing vs explicit}
  \label{fig:explicitvsno}
\end{figure}

\section{Conclusion}
We have presented a method for constructing finally tagless style DSLs with
sharing detection, that allows for DSLs specifying hypergraphs (e.g., functions
with multiple outputs). It also avoids the use of unsafe referencing as
performed when doing observable sharing, c.f. \cite{gill:observablesharing}.

The method has it's drawbacks in terms of memory usage, but these can be
mitigated by explicitly specifying sharing. This does present an extra burden on
the DSL writer to implement explicit sharing when necessary and ensure labels
are not reused. Future work may investigate the use of a preprocessor or plugin
to automate explicit sharing.

>>>>

In the toy example
\begin{minted}{haskell}
class Exp repr where
  variable :: String -> repr Int
  constant :: String -> repr Int
  add :: repr Int -> repr Int -> repr Int
  novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}
the function \mintinline{haskell}{novel} will exhibit an issue in sharing of
it's two outputs when writing it's implementation that cannot be solved using
the methods of \cite{kiselyov:sharing}. We will illustrate this implementation
issue in Section \ref{limitexplicit}. In our work, this translated into the
inability to process large library functions.

@@

We used this language to implement approximations of vector
\mintinline{haskell}{cos} and \mintinline{haskell}{tan} (i.e., a loop body that
computes over an input and output array). The loop body was unrolled by a factor
of four making for a sufficiently large resulting DAG.

We profiled \mintinline{haskell}{cos} using both Kiselyov's method (i.e., hash-consing) and are own
(i.e., trie-consing with some explicit sharing). It wasn't possible to use
Kiselyov's explicit sharing method where meaningful so our code was able to
achieve greater performance, with only a small increase in memory consumption
(see Fig.~\ref{fig:hashvscons}).

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/cos_profile_hashcons_cropped.png}
  \includegraphics[width=0.6\textwidth]{figs/cos_profile_triecons_cropped.png}
  \caption{Comparison of cos using hash-cons vs trie-cons}
  \label{fig:hashvscons}
\end{figure}

For \mintinline{haskell}{tan}, we were unable to generate a resulting code graph using Kiselyov's
method. Using explicit sharing via let constructs wasn't possible due to the
multiple outputs issue and without it the amount of computation due to redundant
hash-consing was simply too large to perform in a reasonable amount of time.

We were able to generate \mintinline{haskell}{tan} using our method both without and with our explicit
sharing, however without explicit sharing we consumed an unreasonable amount of
memory (see Fig.~\ref{fig:explicitvsno}). It should be noted, the main source of
redundant computation in \mintinline{haskell}{tan} is the reuse of computed values of
\mintinline{haskell}{cos} and \mintinline{haskell}{sin}. By simply explicitly
sharing just those values we achieved the significant speedup shown in
Fig.~\ref{fig:explicitvsno}.

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/noexplicit_cropped.png}
  \hfill
  \includegraphics[width=0.6\textwidth]{figs/explicit_cropped.png}
  \caption{Comparison of tan profiling using no explicit sharing vs explicit}
  \label{fig:explicitvsno}
\end{figure}

\section{Conclusion and Future Work}
We have presented a method for constructing finally tagless style DSLs with
sharing detection, that allows for DSLs specifying hypergraphs (e.g., functions
with multiple outputs). It also avoids the use of unsafe referencing as
performed when doing observable sharing, c.f. \cite{gill:observablesharing}.

The method has it's drawbacks in terms of memory usage, but these can be
mitigated by explicitly specifying sharing. This does present an extra burden on
the DSL writer to implement explicit sharing when necessary and ensure labels
are not reused. Future work may investigate the use of a preprocessor or plugin
to automate explicit sharing. Unlike an explicit let construct, it would be
fairly straightforward to automatically bind \mintinline{haskell}{subT}
operations to any DSL function call.
